% Make sure to keep the text within the brackets and the output should turn out correct
\acknowledge{%
First of all, I would like to express my sincere gratitude to my advisor, Professor Sean Meyn, who has been a source of continued wisdom, guidance and support throughout my PhD over the last six years. Right from the time, I was a student in his Stochastic methods class, I was lucky to receive his able mentorship that has made my sailing smooth. With limited prior research experience, it would not have been possible to step into complex arena of stochastic control and Markov chains research, if not for his amazing patience and cheerful personality. I would like to thank Professor Jose Principe, Professor Kamran Mohseni and Professor James Hobert, for agreeing to be part of my supervisory committee and for their critical evaluation of my work. I would also like to thank Professor Eric Moulines for the warm hospitality he bestowed on me during my month long stay at \'{E}cole Polytechnique, Paris in the summer of 2016. This stay opened up avenues for collaborative research and I also got the opportunity to attend the first Data Science Summer School (DS3) in 2017. I would like to thank all the professors at IIT Bombay and College of Engineering, Trivandrum for helping me build a solid foundation, which acted as a springboard for conducting my PhD research. 

I would like to thank my lab mates and colleagues, Adithya Devraj, Joel Mathias, Shuhang Chen, Neil Cammardella, Yue Chen and Surya Dhulipala for helping me with my research and creating a friendly environment to work in. 

When I first landed in Gainesville in Fall 2012, I never imagined it would be my home for the next seven years. After the initial struggle of settling in a new environment, I was made to feel at home by my amazing friends and roommates. I am thankful to Kishore Rajasekar, Shivashankar Halan, Pratyush Chakraborty, Joji Jacob, Sarath Francis, Gayathri Srinivasan, Ravi Venkatraman, Ramya Sivakumar, Yogesh Deshmukh, Thames Harrison for making my life relatively stress-free. I would like to thank Dr. Ravi Ahuja and my team members at Optym for the opportunity to intern there. 

I would like to thank my parents, Lalitha and Radhakrishnan for the immense support, both financial and emotional, that they provided me throughout my life. The freedom of choice that they granted, allowed me to pursue my dream. I am also indebted to my grandparents for their unconditional love and affection. My cousin Dr. Sriram Ganapathy deserves special mention as he was the first person to inspire me to pursue a PhD. I would also like to thank my friend Dr.Krishnakumar Gopalakrishnan for his constant support in matters of research and life in general.

I would like to express my love and gratitude to my wife Dharani Balasubramanian, who has stood by my side for the last four years. This PhD would not have been completed without the tons of emotional support that you gave me to keep me going. Thank you for sticking with me through thick and thin.
 }
